# The SELECT Statement {#select}

```{r echo = FALSE}
# Setting up 
library(DBI)

# Connect to the book's example MySQL database
con <- dbConnect(RMariaDB::MariaDB(), 
             dbname = "farmers_market", 
             host = "localhost", 
             port = 3306,
             user = "root",
             password = keyring::key_get("mysql", "root"))

# No limit on the number of records to be displayed
knitr::opts_knit$set(sql.max.print = NA)
```


## Chapter Notes

The next few chapters introduce basic SQL syntax.

Here's a very simple example:

```{sql connection=con, max.print = 5}
SELECT * FROM farmers_market.product
```

We've selected all columns from the table named "product", and I've used R Markdown's knit options to limit the results displayed to the first five rows. We can use SQL's LIMIT clause for the same effect:

```{sql connection=con}
SELECT * 
FROM farmers_market.product
LIMIT 5
```

Line breaks, tabs and spaces are treated the same by SQL. Usually adding line breaks makes for more legible code.

Instead of asking for all columns, we can specify the ones we want in the SELECT clause:

```{sql connection=con}
SELECT product_id, product_name 
FROM farmers_market.product
LIMIT 5
```

The chapter introduces the ORDER BY clause:

```{sql connection=con}
SELECT product_id, product_name 
FROM farmers_market.product 
ORDER BY product_name
LIMIT 5
```

The default ASC behaviour sorts text alphabetically and numbers from low to high, but we can also use the DESC behaviour:

```{sql connection=con}
SELECT product_id, product_name 
FROM farmers_market.product 
ORDER BY product_name DESC
LIMIT 5
```

We can also sort on multiple columns. The code below sorts first on market date and then on vendor ID:

```{sql connection=con}
SELECT market_date, vendor_id, booth_number 
FROM farmers_market.vendor_booth_assignments 
ORDER BY market_date, vendor_id
LIMIT 5
```


We can also do simple calculations on the columns from within the SELECT clause. The code below contains two calculations like this. First, the columns quantity and cost_to_customer_per_qty are multiplied together. The the *function* ROUND is used to round the resulting price column to 2 digits after the decimal point. The ROUND function also accepts negative numbers to round digits to the left of the decimal point. The AS keyword is used to give the resulting column the *alias* price

```{sql connection=con}
SELECT market_date, customer_id, vendor_id, ROUND(quantity * cost_to_customer_per_qty,2) AS price
FROM farmers_market.customer_purchases
LIMIT 5
```

Another common function is the CONCAT function, which can be used inside the SELECT statement to e.g. combine first and last names into a single column like so: CONCAT(customer_first_name, " ", customer_last_name) AS customer_name. The UPPER function puts text values in all caps.

There is then a section about using the query editor in MySQL workbench to perform some basic checks on your query. I did a little of that but obviously can't display that here. One option that R Markdown allows for though is to assign the result of your query to some R object using the output.var option. Here I've included output.var="transactions" in the SQL chunk options and when I run the query: 

```{sql connection=con, output.var="transactions"}
SELECT market_date, customer_id, vendor_id, ROUND(quantity * cost_to_customer_per_qty, 2) AS price
FROM farmers_market.customer_purchases
```

I get a data frame with 4,221 rows, and I can then use standard R functions to check if the query is returning results as expected. E.g.

```{r}
range(transactions$market_date)
```


## Exercises

### Ex. 1 {-}

#### Question {-}

> Write a query that returns everything in the customer table.

#### Answer {-}

```{sql connection=con}
SELECT *
FROM farmers_market.customer
```


### Ex. 2 {-}

#### Question {-}

> Write a query that displays all of the columns and 10 rows from the customer table, sorted by customer_last_name , then customer_first_name .

#### Answer {-}

```{sql connection=con}
SELECT *
FROM farmers_market.customer
ORDER BY customer_last_name, customer_first_name
LIMIT 10
```


### Ex. 3 {-}

#### Question {-}

> Write a query that lists all customer IDs and first names in the customer table, sorted by first_name.

#### Answer {-}

```{sql connection=con}
SELECT customer_id, customer_first_name
FROM farmers_market.customer
ORDER BY customer_first_name
```




```{r echo = FALSE}
# Disconnecting the database
dbDisconnect(con)
```
