# CASE Statements {#case}

```{r echo = FALSE}
# Setting up 
library(DBI)

# Connect to the book's example MySQL database
con <- dbConnect(RMariaDB::MariaDB(), 
             dbname = "farmers_market", 
             host = "localhost", 
             port = 3306,
             user = "root",
             password = keyring::key_get("mysql", "root"))

# No limit on the number of records to be displayed
knitr::opts_knit$set(sql.max.print = NA)
```


## Chapter Notes

CASE statements are used to create derived columns.

Here's a list of vendors and the kind of items they sell:

```{sql connection=con}
SELECT vendor_id, vendor_name, vendor_type
FROM farmers_market.vendor
```

We can use a CASE statement to add a new column that specifies whether the vendors sell fresh produce or something else:

```{sql connection=con}
SELECT vendor_id, vendor_name, vendor_type, 
  CASE 
    WHEN LOWER(vendor_type) LIKE '%fresh%' 
    THEN 'Fresh Produce' 
    ELSE 'Other' 
  END AS vendor_type_condensed
FROM farmers_market.vendor
```

Note the comma after vendor_type - it looks like CASE is included as part of the SELECT clause. Note also the catch-all ELSE command.

CASE statements are often used to create binary flag fields:

```{sql connection=con}
SELECT market_date, 
  CASE 
    WHEN market_day = 'Saturday' OR market_day = 'Sunday' 
    THEN 1 ELSE 0
  END AS weekend_flag
FROM farmers_market.market_date_info 
LIMIT 5
```

or to bin continuous variables:

```{sql connection=con}
SELECT market_date, customer_id, vendor_id, ROUND(quantity * cost_to_customer_per_qty, 2) AS price, 
  CASE 
    WHEN quantity * cost_to_customer_per_qty < 5.00 
    THEN 'Under $5' 
    WHEN quantity * cost_to_customer_per_qty < 10.00 
    THEN '$5-$9.99' 
    WHEN quantity * cost_to_customer_per_qty < 20.00 
    THEN '$10-$19.99' 
    WHEN quantity * cost_to_customer_per_qty>= 20.00 
    THEN '$20 and Up' 
  END AS price_bin
FROM farmers_market.customer_purchases
LIMIT 10
```



## Exercises

### Ex. 1 {-}

#### Question {-}

> Products can be sold by the individual unit or by bulk measures like lbs. or oz. Write a query that outputs the product_id and product_name columns from the product table, and add a column called prod_qty_type_condensed that displays the word “unit” if the product_qty_type is “unit,” and otherwise displays the word “bulk.”

#### Answer {-}

```{sql connection=con}
SELECT product_id, product_name,
  CASE 
    WHEN LOWER(product_qty_type) LIKE '%unit%' 
    THEN 'unit'
    ELSE 'bulk'
  END AS prod_qty_type_condensed
FROM farmers_market.product
```


### Ex. 2 {-}

#### Question {-}

> We want to flag all of the different types of pepper products that are sold at the market. Add a column to the previous query called pepper_flag that outputs a 1 if the product_name contains the word “pepper” (regardless of capitalization), and otherwise outputs 0.

#### Answer {-}

```{sql connection=con}
SELECT product_id, product_name,
  CASE 
    WHEN LOWER(product_qty_type) LIKE '%unit%' 
    THEN 'unit'
    ELSE 'bulk'
  END AS prod_qty_type_condensed,
  CASE 
    WHEN LOWER(product_name) LIKE '%pepper%' 
    THEN 1
    ELSE 0
  END AS pepper_flag  
FROM farmers_market.product
```

### Ex. 3 {-}

#### Question {-}

> Can you think of a situation when a pepper product might not get flagged as a pepper product using the code from the previous exercise?

#### Answer {-}

The product could just be called "Jalapenos - Organic" or similar, anything without the word pepper in it won't be flagged.



```{r echo = FALSE}
# Disconnecting the database
dbDisconnect(con)
```
