# Building SQL Datasets for Analytical Reporting {#building_datasets}

```{r echo = FALSE}
# Setting up 
library(DBI)

# Connect to the book's example MySQL database
con <- dbConnect(RMariaDB::MariaDB(), 
             dbname = "farmers_market", 
             host = "localhost", 
             port = 3306,
             user = "root",
             password = keyring::key_get("mysql", "root"))

# No limit on the number of records to be displayed
knitr::opts_knit$set(sql.max.print = NA)
```


## Chapter Notes

This chapter contains some helpful advice about creating a data sets to answer questions of interest, particularly when you want that data set to be reusable. Here are some tips:

* Consider the level of granularity required to answer the business question - we don't want to have to aggregate each time we query the data set, but we also don't want to have to create ad hoc queries at a lower level of granularity too often. What is the lowest level of granularity that is going to regularly be useful to answer the questions your data set is built for?
* Use human readable columns where possible (e.g. vendor name instead of ID)
* Include useful dimensions for grouping and slicing (e.g. product category)
* The first question you are asked will likely not be the only one. Try to anticipate follow-up questions and build your data set to be able to answer them.

The chapter goes on to introduce two new ways of storing queries and the results of queries for reuse: common table expressions (CTEs), and views. Another approach, creating new database tables, will be introduced in a later chapter.

CTEs or WITH clauses allow you to create an alias for a query, which you can then reference in other queries in the same way you would a table.

The way that this works is that you write a WITH clause, assigning one or more queries an alias, and then below the WITH clause you can start your SELECT query and reference the alias(es) like you would a table. In the WITH clause, each query can reference the alias of any previous queries in the clause. Here's the syntax:

```{sql connection=con, eval = FALSE}

WITH [query_alias] AS -- first query alias
( 
        [query] -- first query definition
        ),
        
    [query_2_alias] AS -- second query alias 
        ( 
        [query_2] -- second query definition 
) 

SELECT [column list] -- here's your regular SELECT query
FROM [query_alias] -- you can reference the aliases just defined

```

Here's the example from the chapter:

```{sql connection=con}
WITH sales_by_day_vendor AS (
  SELECT cp.market_date, md.market_day, md.market_week, md.market_year, 
          cp.vendor_id, v.vendor_name, v.vendor_type, 
          ROUND(SUM(cp.quantity * cp.cost_to_customer_per_qty),2) AS sales 
  FROM farmers_market.customer_purchases AS cp 
  LEFT JOIN farmers_market.market_date_info AS md 
      ON cp.market_date = md.market_date 
  LEFT JOIN farmers_market.vendor AS v 
      ON cp.vendor_id = v.vendor_id 
  GROUP BY cp.market_date, cp.vendor_id 
  ORDER BY cp.market_date, cp.vendor_id
) 


SELECT s.market_year, s.market_week, 
        SUM(s.sales) AS weekly_sales
FROM sales_by_day_vendor AS s        -- referencing CTE defined above
GROUP BY s.market_year, s.market_week
LIMIT 10
```

CTEs use cleaner-looking code and so are easier to understand than a sub-query.

Another approach is to create a database *view*. Here's the chapter:

> A view is treated just like a table in SQL, the only difference being that it has run when it's referenced to dynamically generate a result set (where a table stores the data instead of storing the query), so queries that reference views can take longer to run than
queries that reference tables.

You'll need database permissions to create a view. If you have the permissions you can create a view by preceding your SELECT statement with

```{sql connection=con, eval = FALSE}
CREATE VIEW [schema_name].[view_name] AS
```

Here's an example from the chapter:

```{sql connection=con, eval = FALSE}
CREATE VIEW farmers_market.vw_sales_by_day_vendor AS 
  SELECT cp.market_date, md.market_day, md.market_week, md.market_year, 
          cp.vendor_id, v.vendor_name, v.vendor_type, 
          ROUND(SUM(cp.quantity * cp.cost_to_customer_per_qty),2) AS sales
  FROM farmers_market.customer_purchases AS cp 
  LEFT JOIN farmers_market.market_date_info AS md 
    ON cp.market_date = md.market_date 
  LEFT JOIN farmers_market.vendor AS v 
    ON cp.vendor_id = v.vendor_id
GROUP BY cp.market_date, cp.vendor_id
ORDER BY cp.market_date, cp.vendor_id
```

Here the view has been named with a preceding "vw_" to make it clear that this a view when writing queries that reference it.

You can then use that stored view like you would a table:

```{sql connection=con}
SELECT * 
FROM farmers_market.vw_sales_by_day_vendor AS s -- referencing created view
WHERE s.market_date BETWEEN '2020-04-01' AND '2020-04-30'
      AND s.vendor_id = 7
ORDER BY market_date
```

Also remember that the view is dynamic - it updates each time it is run and so will reflect any changes in the database.


## Exercises

### Ex. 1 {-}

#### Question {-}


#### Answer {-}

## Further Reading {-}



```{r echo = FALSE}
# Disconnecting the database
dbDisconnect(con)
```
