--- 
title: "Notes on 'SQL for Data Scientists' by Renée M. P. Teate"
author: "Jake Lawler"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: jake-lawler/SQLforDataScientists
description: "My Notes on 'SQL for Data Scientists'."
---

# Preface {-}

```{r echo = FALSE, eval = FALSE}

# Create a live view of the book.

library(bookdown)
serve_book(dir = ".", output_dir = "_book", preview = TRUE, in_session = TRUE, quiet = FALSE)

```

These are my notes on 'SQL for Data Scientists' by Renée M. P. Teate. I've only just started so there's not much here yet.



Here are the sources I referred to to get MySQL up-and-running from within R Markdown:

* The answer from 'Harley' [here](https://stackoverflow.com/questions/50544230/connecting-to-mysql-from-r) on connecting to MySQL.

* The [reference page](https://rmariadb.r-dbi.org/reference/index.html) for RMariaDB.

* The information on including SQL in R Markdown from Yihui Xie's [R Markdown book](https://bookdown.org/yihui/rmarkdown/language-engines.html#sql).

Here's an example of the code I'm using to connect to the book's example database, I'll hide this in the future chapters:

```{r}
# Setting up 
library(DBI)

# Connect to the book's example MySQL database
con <- dbConnect(RMariaDB::MariaDB(), 
             dbname = "farmers_market", 
             host = "localhost", 
             port = 3306,
             user = "root",
             password = keyring::key_get("mysql", "root"))

# No limit on the number of records to be displayed
knitr::opts_knit$set(sql.max.print = NA)
```


You can then run queries by using R Markdown's support for SQL code snippets:

```{sql connection=con, max.print = 10}
SELECT * FROM farmers_market.product
```


Then disconnecting the database at the end of each chapter:

```{r}
# Disconnecting the database
dbDisconnect(con)
```


